Model Finding for Recursive Functions in SMT

### 2016-01-26 ijcar2016/body.tex
Fixed numbers from misclassified benchmarks
removed refs to case studies

\newcommand\innerkw[1]{\textrm{#1}}

@@

%\newcommand{\functerms}{\mathbf{DFT}}
\newcommand{\vars}{\mathbf{V}}
\newcommand{\M}{\mathcalx{M}}
\newcommand{\I}{\mathcalx{I}}
\newcommand{\J}{\mathcalx{J}}
\def\AIF{\qtab\keyword{if}\ }
\def\THEN{\ \keyword{then}\ }
\def\AELSE{\untab\qtab\keyword{else}\ }
\def\FI{\untab}
\def\RETURN{\keyword{return}\ }
\def\ENDPROC{\untab}
\newcommand{\conv}{\mathcalx{A}}

@@

The approach is implemented as a preprocessor in \cvc and in the Nunchaku
model finder (Section~\ref{sec:implementations}). As a case study, we employed
Nunchaku on Isabelle/HOL~\cite{nipkow-et-al-2002} formalizations of binary
trees and coinductive lists (Section~\ref{sec:case-studies}).
%
We evaluated the two evaluation on benchmarks from
IsaPlanner~\cite{DBLP:conf/itp/JohanssonDB10}, Leon~\cite{blanc2013overview},
and Isabelle/HOL, to demonstrate that this translation improves the
effectiveness of the SMT solvers \cvc and \ziii for finding countermodels to
verification conditions (Section~\ref{sec:evaluation}). Unlike earlier work,
our approach can be combined with off-the-shelf SMT solvers
(Section~\ref{sec:related-work}).

@@

Nunchaku is a new higher-order model finder designed to be integrated with
several proof assistants. The first version, 0.1, was released in January 2016
with support for (co)algebraic datatypes, (co)recursive functions and
(co)inductive predicates. Support for higher-order functions is planned for
the next release. We have developed a preliminary Isabelle frontend and are
planning further frontends for Coq, the TLA\textsuperscript{+} Proof System,
and possibly other systems.

@@

\section{Case Studies}
\label{sec:case-studies}

\subsection{Lazy Lists}

XXX

\subsection{Huffman Coding}

XXX

>>>>

\newcommand\innerkw[1]{\textrm{#1}}
\newcommand\keyw[1]{\texttt{#1}}

@@

%\newcommand{\functerms}{\mathbf{DFT}}
\newcommand{\vars}{\mathbf{V}}
\newcommand{\M}{\mathcalx{M}}
\newcommand{\I}{\mathcalx{I}}
\newcommand{\J}{\mathcalx{J}}
\newcommand{\conv}{\mathcalx{A}}

@@

The approach is implemented as a preprocessor in \cvc and in the Nunchaku
model finder (Section~\ref{sec:implementations}).
%As a case study, we employed
%Nunchaku on Isabelle/HOL~\cite{nipkow-et-al-2002} formalizations of
%coinductive lists and binary trees (Section~\ref{sec:case-studies}).
%
We evaluated the two evaluation on benchmarks from
IsaPlanner~\cite{DBLP:conf/itp/JohanssonDB10}, Leon~\cite{blanc2013overview},
and Isabelle/HOL, to demonstrate that this translation improves the
effectiveness of the SMT solvers \cvc and \ziii for finding countermodels to
verification conditions (Section~\ref{sec:evaluation}). Unlike earlier work,
our approach can be combined with off-the-shelf SMT solvers
(Section~\ref{sec:related-work}).

An earlier version of this paper was presented at the SMT 2015 workshop in San
Francisco \cite{reynolds-et-al-2015-smt}. This paper extends the workshop
paper with proof sketches, an expanded implementation section covering
Nunchaku and relevant \cvc optimizations, %the case studies,
and the evaluation on Isabelle benchmarks produced by Nunchaku.

@@

Nunchaku is a new higher-order model finder designed to be integrated with
several proof assistants. The first version, 0.1, was released in January 2016
with support for (co)algebraic datatypes, (co)recursive functions and
(co)inductive predicates. Support for higher-order functions is planned for
the next release. We have developed a preliminary Isabelle frontend and are
planning further frontends for Coq, the TLA\textsuperscript{+} Proof System,
and possibly other proof assistants.

@@

%\section{Case Studies}
%\label{sec:case-studies}
%We show Nunchaku in action on two small Isabelle formalizations:\ coinductive
%lists and a binary data structure called AA trees. These two examples were
%used before to demonstrate Nitpick \cite{}.
%\subsection{Coinductive Lists}
%TODO: Adapt!
%\newcommand\List[1]{[#1]}
%\newcommand\Nil{\List{\vthinspace}}
%\newcommand\bool{\mathit{bool}}
%The Isabelle codatatype $\alpha\>\mathit{llist}$ of lazy lists is generated by the
%constructors $\const{LNil} : {\alpha\>\mathit{llist}}$ and $\const{LCons} : {\alpha\to
%\alpha\>\mathit{llist} \to \alpha\>\mathit{llist}}$. It is of particular interest to countermodel finding
%because many %basic
%properties of finite lists do not carry over to infinite
%lists, often in baffling ways. To illustrate this, we conjecture that
%appending $\textit{ys}$ to $\textit{xs}$ yields $\textit{xs}$ iff
%$\textit{ys}$ is $\const{LNil}$:
%\[\textit{xs} \mathbin{@} \textit{ys} = \textit{xs} \,\longleftrightarrow\,
%  \textit{ys} = \const{LNil}\]
%The operator $\mathbin{@}$ is defined by primitive corecursion by the equation
%\[\textit{xs} \mathbin{@} \textit{ys} =
%  (\innerkw{case}~\textit{xs}~\innerkw{of}~\const{LNil} \Rightarrow \textit{ys}
%  \mid \const{LCons}\;x\;\textit{xs}' \Rightarrow \const{LCons}\;x\;(\textit{xs}' \mathbin{@} \textit{ys}))\]
%Nunchaku immediately
%finds the countermodel $\textit{xs} = \textit{ys} =
%\List{0, 0, \ldots}$, in which a cardinality of 1 is sufficient for
%$\alpha$ and $\alpha\>\mathit{llist}$, and the bisimilarity predicate $\sim$ is unrolled only
%once. Indeed, appending $\List{0, 0, \ldots} \not= \Nil{}$ to $\List{0, 0, \ldots}$ leaves
%$\List{0, 0, \ldots}$ unchanged. Many other counterexamples are possible---for example,
%$\textit{xs} = \List{0, 0, \ldots}$ and $\textit{ys} = \List{1}$---but Nitpick
%tends to reuse the objects that are part of a subterm-closed substructure to
%keep cardinalities low. Although very simple, the counterexample is beyond
%Quickcheck's and Refute's reach, since they do not support codatatypes.
%The next example requires the following lexicographic order predicate:
%%
%\begin{quote}
%$\keyw{coinductive}~\vthinspace\smash{{\preceq}\typ{\mathit{nat}\>\mathit{llist}\to\mathit{nat}\>\mathit{llist}\to\bool}}~\keyw{where}$ \\
%$\const{LNil} \preceq \textit{xs}$ \\
%$x \le y \,\Longrightarrow\, \const{LCons}~x~\textit{xs} \preceq \const{LCons}~y~\textit{ys}$ \\
%$\textit{xs} \preceq \textit{ys} \,\Longrightarrow\, \const{LCons}~x~\textit{xs} \preceq \const{LCons}~x~\textit{ys}$
%\end{quote}
%%
%The intention is to define a linear order on lazy lists of
%natural numbers, and hence the following properties should hold:
%%
%\begin{align*}
%\textsc{Refl}{:}\kern.75em
%  & \textit{xs} \preceq \textit{xs}
%& \textsc{Antisym}{:}\kern.75em
%  & \textit{xs} \preceq \textit{ys} \mathrel{\land} \textit{ys} \preceq \textit{xs} \longrightarrow \textit{xs} = \textit{ys} \\
%\textsc{Linear}{:}\kern.75em
%  & \textit{xs} \preceq \textit{ys} \mathrel{\lor} \textit{ys} \preceq \textit{xs}
%& \textsc{Trans}{:}\kern.75em
%  & \textit{xs} \preceq \textit{ys} \mathrel{\land} \textit{ys} \preceq \textit{zs} \longrightarrow \textit{xs} \preceq \textit{zs}%.
%\end{align*}
%%
%However, Nitpick finds a counterexample for \textsc{Antisym}: $\textit{xs} = \List{1,
%1}$ and $\textit{ys} = \List{1}$. On closer inspection, the assumption $x \leq y$ of
%the second introduction rule for $\preceq$ should have been $x < y$; otherwise,
%any two lists \textit{xs}, \textit{ys} with the same head satisfy
%$\textit{xs} \preceq \textit{ys}$. Once we repair the specification, no more
%counterexamples are found for the four properties up to cardinality 6
%for $\textit{nat}$ and $\mathit{nat}\>\mathit{llist}$ within the time limit of 30~seconds.
%%This is a strong indication that the properties hold.
%Andreas Lochbihler used Isabelle
%to prove all four properties \cite{lochbihler-2010-coinduction}.
%We could continue like this and sketch a full-blown theory of lazy lists. Once the
%definitions and main theorems %are in place and
%have been thoroughly tested using
%Nitpick, we could start working on the proofs. Developing theories this way
%can save a lot of time, because faulty theorems and definitions are discovered
%early. %in the process.
%\subsection{AA Trees}
